---
title: "Bayesian 4"
author: "Let's run this"
---

## Model: Wallcreeper


Birds that have declined in the last years

We will look at (simulate) data from 1989 to 2005

We sample random quadrats each year and record (proportion of quadrats)

We will use a normal linear model for this

## Simulate our population

$$
y_i = \beta_0 + \beta_1 year_i + \epsilon_i 
$$

$$
y_i = \alpha + \beta \times year_i + \epsilon_i 
$$

```{r}

n = 16 (years)
a = 40 (intercept)
b = -0.5 (slope)
sigma = 5 
sigma2 = sigma^2
x = 1:16

```
y is proportion of quadrats
Save true values in an object

```{r}
truth <- c(alpha = a, beta = b, sigma = sqrt(sigma2))
```


## Simulation

```{r}
set.seed(12320)
```

```{r}
y<- a+b*x + rnorm(n,0,sigma)

```

## Plot the data

CHANGE THE X AXIS!

```{r}


```



## Run the model in frequentist
```{r}
model <- lm(y ~ x)
summary(model) 
lm_est <- c(coef(model), sigma = sigma(model)) 
 lm_est 

```

## Run the stan model

Download the stan package

```{r}
library(rstan)
```

## Let's run the model

First step: Bundle the data:

Use a list to put response variable, explanatory variable and n(number of observations)

```{r}
dataList <- list(y = y, x = x, n = n)
```

## Step 2 (do not run)
Let's break this up

```{r eval=FALSE}
# Write text file with model description in Stan language
cat(file = "model.stan", # This line is R code
"data { // This is the first line of Stan code
  int <lower = 0> n; // Define the format of all data
  vector[n] y; //. . . including the dimension of vectors
  vector[n] x; //
}
parameters { // Define format for all parameters
  real alpha;
  real beta;
  real <lower = 0> sigma; // sigma (sd) cannot be negative
}
transformed parameters {
  vector[n] mu;
  for (i in 1:n){
    mu[i] = alpha + beta * x[i]; // Calculate linear predictor
  }
}
model {
  // Priors
  alpha ~ normal(0, 100);
  beta ~ normal(0, 100);
  sigma ~ cauchy(0, 10);
  // Likelihood (could be vectorized to increase speed)
  for (i in 1:n){
    y[i] ~ normal(mu[i], sigma);
  }
}

" )

```


## Data and parameters

```{r eval=FALSE}
cat(file = "model.stan", # This line is R code
"data { // This is the first line of Stan code
  int <lower = 0> n; // Define the format of all data
  vector[n] y; //. . . including the dimension of vectors
  vector[n] x; //
}

parameters { // Define format for all parameters
  real alpha;
  real beta;
  real <lower = 0> sigma; // sigma (sd) cannot be negative
}
```

## Transformed parameter
```{r}
transformed parameters {
  vector[n] mu;
  for (i in 1:n){
    mu[i] = alpha + beta * x[i]; // Calculate linear predictor
  }

```

## Model
```{r eval=FALSE}
model {
  // Priors
  alpha ~ normal(0, 100);
  beta ~ normal(0, 100);
  sigma ~ cauchy(0, 10);
  // Likelihood
  for (i in 1:n){
    y[i] ~ normal(mu[i], sigma);
  }
}
```

## All code

```{r}
# Write text file with model description in Stan language
cat(file = "model.stan", # This line is R code
"data { // This is the first line of Stan code
  int <lower = 0> n; // Define the format of all data
  vector[n] y; //. . . including the dimension of vectors
  vector[n] x; //
}
parameters { // Define format for all parameters
  real alpha;
  real beta;
  real <lower = 0> sigma; // sigma (sd) cannot be negative
}
transformed parameters {
  vector[n] mu;
  for (i in 1:n){
    mu[i] = alpha + beta * x[i]; // Calculate linear predictor
  }
}
model {
  // Priors
  alpha ~ normal(0, 100);
  beta ~ normal(0, 100);
  sigma ~ cauchy(0, 10);
  // Likelihood (could be vectorized to increase speed)
  for (i in 1:n){
    y[i] ~ normal(mu[i], sigma);
  }
}

" )

```
## Next step 
Set settings and run it

```{r}
# HMC settings
ni <- 3000 ; nb <- 1000 ; nc <- 4 ; nt <- 1
# Call STAN 

modelstan <- stan(file = "model.stan", data = dataList,
chains = nc, iter = ni, warmup = nb, thin = nt)


```
