---
title: "Bayesian statistics: part two"
author: "SNR 610"
execute: 
  echo: true
format:
  revealjs:
    incremental: true   
    scrollable: true
    smaller: true
    chalkboard: true
    mermaid:
      theme: default
    fontsize: 62px
---

## Bayesian

A different statistical framework

We have used frequentist so far

| Topic | Frequentist | Bayesian |
|------------------------|------------------------|------------------------|
| Parameters | Fixed unknown | Random variable |
| Data | Random | Fixed |
| Inference | Based on a sampling distribution | Based on posterior |
| Output | P-value, Confidence intervals, betas | Credible intervals and posterior distribution |
| Computational needs | low | high |

-   Data, random? Yes...

-   $y_i = \beta_0 + \beta_1 + \epsilon_i$

## Bayesian take away

Different way of looking at things... we will explore how

## Bayes Theorem

$$
\Huge P(\theta |data) = \frac{P(data|\theta) \times P(\theta)}{P(data)}
$$

-   $P(\theta|data)$ : Posterior, probability of theta, given the data
-   $P(data|\theta)$ : Likelihood! $\mathcal{L}(\theta|data) = P(data|\theta)$
-   $P(\theta)$: Prior (who knows what this is!)
-   $P(data)$: Marginal (who knows!)

## Challenges

-   Challenge 1:

    -   You go out to a Forest that has been separated in multiple plots. You are wondering what the count of total trees per plot is ($\lambda$). You only can count the number of trees in one plot chosen at random. You count 37 trees, Estimate the Likelihood of different values of $\lambda$ as well as the maximum likelihood.

-   Challenge 2:

    -   You go to the same Forest, but were able to count three plots. They had 37, 41 and 35 trees. Estimate the Likelihood of different values of $\lambda$

## Answers Q1

```{r}
x=seq(25,50, by=0.25)
y=dpois(37,x)
plot(x,y,type="l", col="blue", lwd=2,ylab="Likelihood", xlab="Lambda")
```

## Answers Q2

```{r}
x=seq(25,50, by=0.25)
y=dpois(37,x)*dpois(41,x)*dpois(35,x)
plot(x,y,type="l", col="blue", lwd=2,ylab="Likelihood", xlab="Lambda")
```

## Bayes Theorem

$$ \Huge P(\theta |data) = \frac{P(data|\theta) \times P(\theta)}{P(data)} $$

-   $P(\theta|data)$ : Posterior, probability of theta, given the data
-   $P(data|\theta)$ : Likelihood! $\mathcal{L}(\theta|data) = P(data|\theta)$
-   $P(\theta)$: Prior (who knows what this is!)
-   $P(data)$: Marginal (who knows!)

## Prior

Before talking about priors, I want to tell you all a story

##