---
title: "Bayesian statistics"
author: "SNR 610"
execute: 
  echo: true
format:
  revealjs:
    theme: white                      
    incremental: true   
    scrollable: true
    smaller: true
    chalkboard: true
    fontsize: 55px
revealjs-plugins:
  - revealjs-text-resizer
---

## Reminder

-   What is probability?

-   Certainty that an event will happen

-   Example:

    -   X = number of heads in 3 tosses

    -   x = exactly 2 heads

    -   $\theta=0.5$ (probability)

    -   $P(X=x | \theta) = 3/8$

    -   $P(\text{Number of heads in 3 tosses}= 2 | 0.5) = 3/8$

## Why?

```{mermaid}

flowchart LR
  A[HEADS] --> B[HEADS]
  A --> C[TAILS]
  B --> D[HEADS]
  B --> E[TAILS]
  C --> F[HEADS]
  C --> G[TAILS]
 

```

```{mermaid}

flowchart LR
  A[TAILS] --> B[HEADS]
  A --> C[TAILS]
  B --> D[HEADS]
  B --> E[TAILS]
  C --> F[HEADS]
  C --> G[TAILS]
 

```

## Probability

-   Probability of 1 head?

-   Probability of 3 heads?

-   Probability of 0 heads?

-   We treat $\theta$ as fixed

## Probability

```{r}
dbinom(2,3,0.5)
```

```{r}
dbinom(1,3,0.5)
```

## Probability

```{r}
dbinom(0:3,3,0.5)
```

```{r echo=F}
barplot(dbinom(0:3,3,0.5), names.arg = 0:3, main="number of heads")
```

## Probability

```{r echo=F}
barplot(dbinom(0:100,100,0.5), names.arg = 0:100, main="number of heads")
```

## Question

You have an unfair coin. What is the probability that you will get 40 heads in 100 tosses?
