---
title: "Bayesian 4"
author: "SNR 610"
execute: 
  echo: true
format:
  revealjs:
    incremental: true   
    scrollable: true
    smaller: true
    chalkboard: true
    mermaid:
      theme: default
    fontsize: 62px
---

## Model: Wallcreeper

```{r echo=FALSE}
set.seed(5480)

```

-   Birds that have declined in the last years

-   We will look at (simulate) data from 1989 to 2005

-   We sample random quadrats each year and record (proportion of quadrats)

-   We will use a normal linear model for this

## Simulate our population

$$
y_i = \beta_0 + \beta_1 year_i + \epsilon_i 
$$

$$
y_i = \alpha + \beta \times year_i + \epsilon_i 
$$

```{r}

n = 16 #(years)
a = 40 #(intercept)
b = -0.7 #(slope)
sigma = 4
sigma2 = sigma^2
x = 1:16

```

y is proportion of quadrats

Save true values in an object

```{r}
truth <- c(alpha = a, beta = b, sigma = sigma)
```

## Simulation

```{r}
y<- a+b*x + rnorm(n,0,sigma)

```

## Plot the data {.nostretch}

CHANGE THE X AXIS!

```{r}
library(ggplot2)

df<-data.frame(x=x,y=y)
ggplot(df,aes(x=x+1989,y=y))+
  geom_point(size=2.5)+
  ylab("Prop. occupied (%)")+
  xlab("Year")+theme_classic()

```

## Run the model in frequentist {.nostretch}

```{r}
model <- lm(y ~ x)
summary(model) 
lm_est <- c(coef(model), sigma = sigma(model)) 
 lm_est 

```

```{r echo=FALSE}
ggplot(df,aes(x=x+1989,y=y))+
  geom_point(size=1.5)+
  geom_smooth(method = lm)+
  ylab("Prop. occupied (%)")+
  xlab("Year")+theme_classic()

```

## Run the stan model

Download the stan package

```{r}
library(rstan)
```

## Let's run the model

First step: Bundle the data:

Use a list to put response variable, explanatory variable and n(number of observations)

```{r}
dataList <- list(y = y, x = x, n = n)
```

## Step 2 (do not run)

Let's break this up

```{r eval=FALSE}
# Write text file with model description in Stan language
cat(file = "model.stan", # This line is R code
"data { // This is the first line of Stan code
  int <lower = 0> n; // Define the format of all data
  vector[n] y; //. . . including the dimension of vectors
  vector[n] x; //
}
parameters { // Define format for all parameters
  real alpha;
  real beta;
  real <lower = 0> sigma; // sigma (sd) cannot be negative
}
transformed parameters {
  vector[n] mu;
  for (i in 1:n){
    mu[i] = alpha + beta * x[i]; // Calculate linear predictor
  }
}
model {
  // Priors
  alpha ~ normal(0, 100);
  beta ~ normal(0, 100);
  sigma ~ cauchy(0, 10);
  // Likelihood (could be vectorized to increase speed)
  for (i in 1:n){
    y[i] ~ normal(mu[i], sigma);
  }
}

" )

```

## Data and parameters

```{r eval=FALSE}
cat(file = "model.stan", # This line is R code
"data { // This is the first line of Stan code
  int <lower = 0> n; // Define the format of all data
  vector[n] y; //. . . including the dimension of vectors
  vector[n] x; //
}

parameters { // Define format for all parameters
  real alpha;
  real beta;
  real <lower = 0> sigma; // sigma (sd) cannot be negative
}
```

## Transformed parameter

```{r eval=FALSE}
transformed parameters {
  vector[n] mu;
  for (i in 1:n){
    mu[i] = alpha + beta * x[i]; // Calculate linear predictor
  }

```

## Model

```{r eval=FALSE}
model {
  // Priors
  alpha ~ normal(0, 100);
  beta ~ normal(0, 100);
  sigma ~ cauchy(0, 10);
  // Likelihood
  for (i in 1:n){
    y[i] ~ normal(mu[i], sigma);
  }
}
```

## All code

```{r}
# Write text file with model description in Stan language
cat(file = "model.stan", # This line is R code
"data { // This is the first line of Stan code
  int <lower = 0> n; // Define the format of all data
  vector[n] y; //. . . including the dimension of vectors
  vector[n] x; //
}
parameters { // Define format for all parameters
  real alpha;
  real beta;
  real <lower = 0> sigma; // sigma (sd) cannot be negative
}
transformed parameters {
  vector[n] mu;
  for (i in 1:n){
    mu[i] = alpha + beta * x[i]; // Calculate linear predictor
  }
}
model {
  // Priors
  alpha ~ normal(0, 100);
  beta ~ normal(0, 100);
  sigma ~ cauchy(0, 10);
  // Likelihood (could be vectorized to increase speed)
  for (i in 1:n){
    y[i] ~ normal(mu[i], sigma);
  }
}

" )

```

## Next step

Set settings and run it

```{r}
# HMC settings
ni <- 3000 ; nb <- 1000 ; nc <- 4 ; nt <- 1
# Call STAN 

modelstan <- stan(file = "model.stan", data = dataList,
chains = nc, iter = ni, warmup = nb, thin = nt)


```

## Get results

```{r}
rstan::traceplot(modelstan)
```

## Get results

```{r}
print(modelstan, dig = 2)
```

## Get results

```{r}
stan_est <- summary(modelstan)$summary[1:3,1]

```

```{r}
comp <- cbind(truth = truth, lm = lm_est, Stan = stan_est)
print(comp, 4)

```

## Challenge

-   Repeat exercise, but with males and females

-   <https://rpubs.com/amolina/bayesexample>

-   Also you could try, run one of the models (e.g., GLM) run in class in Bayesian
