---
title: "Bayesian statistics"
author: "SNR 610"
execute: 
  echo: true
format:
  revealjs:
    incremental: true   
    scrollable: true
    smaller: true
    chalkboard: true
    mermaid:
      theme: default
    fontsize: 62px
---

## Today

Probability and Likelihood (step 1 out of 3)

## Reminder

-   What is probability?

-   Certainty that an event will happen

-   Example:

    -   X = number of heads in 3 tosses

    -   x = exactly 2 heads

    -   $\theta=0.5$ (probability)

    -   $P(X=x | \theta) = 3/8$

    -   $P(\text{Number of heads in 3 tosses}= 2 | 0.5) = 3/8$

## Why?

```{mermaid}
%%| echo: false

flowchart LR
  A[HEADS] --> B[HEADS]
  A --> C[TAILS]
  B --> D[HEADS]
  B --> E[TAILS]
  C --> F[HEADS]
  C --> G[TAILS]
```

```{mermaid}
%%| echo: false


flowchart LR
  A[TAILS] --> B[HEADS]
  A --> C[TAILS]
  B --> D[HEADS]
  B --> E[TAILS]
  C --> F[HEADS]
  C --> G[TAILS]
 
```

## Probability

-   3 tosses

-   Probability of 1 head?

-   Probability of 3 heads?

-   Probability of 0 heads?

-   We treat $\theta$ as fixed (0.5)

## Probability

We can actually estimate any probability as long as we know the parameters

-   Binomial distribution parameters: n and p

-   n is number of trials

-   $P(X=x | \theta) = Y$

-   $P(successes = x | \text{p and n}) = Y$

-   $P(successes = 12 | \text{0.5 and 50}) = 0.0001078$

-   $P(successes \ge  57 | \text{0.5 and 100}) = 0.0966$

-   We use `dbinom`. Like this: `dbinom(x,n,p)` Example: `dbinom(12,50,0.5)`

-   For $\ge$ 57: `sum(dbinom(57:n,n,p))`

## Probability

```{r}
dbinom(2,3,0.5)
```

```{r}
dbinom(1,3,0.5)
```

## Probability

```{r}
dbinom(0:3,3,0.5)
```

```{r echo=F}
barplot(dbinom(0:3,3,0.5), names.arg = 0:3, main="number of heads")
```

## Probability

```{r echo=F}
barplot(dbinom(0:100,100,0.5), names.arg = 0:100, main="number of heads")
```

```{r eval=FALSE}
dbinom(0:100,100,0.5)
```

## Parameters of binomial

Probability and n!

## Parameters of a Normal distribution

Who can remember them?

-   Mean and sd (or variance)

-   For a population of males in the USA, these are the parameters:

    -   $\mu = 177.6cm$

    -   $\sigma = 9.7cm$

```{r echo=FALSE}
x <- seq(140, 220, 1)
data<-dnorm(x, 177.6, 9.7)
plot(NULL, xlim=c(140,220), ylim=c(0, max(data)),  ylab = 'Probability', xlab = 'Height',
     main = "Distribution of Height for American Males between 20 and 28" )
curve(dnorm(x, 177.6, 9.7), from = 140, to = 220, lwd = 4,
      col = 'steelblue',add=T)
abline(v=177.6)
```

-   What is the probability of sampling a random man with a heigh of 172.85431 cm?

-   What is the probability of sampling one random man and him being between 170 and 171 cm?

```{r}
pnorm(171,177.7,9.7)-pnorm(170,177.7,9.7)
```

```{r echo=FALSE}
x <- seq(140, 220, 1)
data<-dnorm(x, 177.6, 9.7)
plot(NULL, xlim=c(140,220), ylim=c(0, max(data)),  ylab = 'Probability', xlab = 'Height',
     main = "Distribution of Height for American Males between 20 and 28" )
curve(dnorm(x, 177.6, 9.7), from = 140, to = 220, lwd = 4,
      col = 'steelblue',add=T)
abline(v=171, lwd=2, lty=2, col="salmon")
abline(v=172, lwd=2, lty=2, col="salmon")

```

## Normal Distribution

By knowing the mean and sd we can plot any normal distibution

We can estimate any probability as well

Probability of observing X given the parameters

-   $P ( 170 \le X \le 171) | \mu = 177.6, \sigma = 9.7)$

## Poisson Distribution

-   One parameter

-   $\lambda$

-   Represents both mean and sd

-   When $\lambda = 7$ what is the probability of seeing 5 hummingbirds visit a flower?

-   $P(X=5|\lambda =7)$

## Poisson Distribution

```{r}
dpois(5,7)
```

```{r}
barplot(dpois(0:20,7), names.arg = 0:20, main="number of hummingbirds")
```

## Takeaway

Probability: Chance of an event occuring

-   Chance of data being something, given the parameters

-   The parameters do not change. For example, for a p of 0.5 we can estimate the prob of 0, 1, 2... 100 heads out of 100

## Question

Test your knowledge...

-   You have a coin that you do not know if it is fair or unfair. What is the probability that you will get 40 heads in 100 tosses?

## Probability

We cannot estimate it if we do not have the parameters!

-   We usually do not have them

-   It is what we are trying to estimate!!!!

-   If we had them, there would be no point on sampling/experiments

-   Examples:

    -   Probability of infection (infection rate)

    -   Yield of a crop

    -   Annual sales

    -   community composition

## What do we have?

Data!

## Unfair or fair coin

You are smart, so you knew we couldn't estimate it! You decide to grab the coin, and do 100 flips.

-   You get 65 heads and 35 tails

-   What value of P (prob of heads) is more likely:

    -   0.3

    -   0.5

    -   0.7

## We can estimate the Likelihood

Likelihood of a parameter given our data

-   Probability: $P(X=x|\theta)$

-   x = data and $\theta$ = parameter

-   Likelihood: $\mathcal{L}(x|\theta = y)$

## In other words

-   Likelihood: $\mathcal{L}(65|p= 0.3 \; \& \; n=100)$

-   Likelihood: $\mathcal{L}(65|p= 0.5 \; \& \; n=100)$

-   Likelihood: $\mathcal{L}(65|p= 0.7 \; \& \; n=100)$

## In other words

Likelihood: $\mathcal{L}(65|p= 0.3 \; \& \; n=100)$

Likelihood: $\mathcal{L}(65|p= 0.5 \; \& \; n=100)$

Likelihood: $\mathcal{L}(65|p= 0.7 \; \& \; n=100)$

```{r}
dbinom(65,100,0.3)  
 

```

## In other words

Likelihood: $\mathcal{L}(65|p= 0.3 \; \& \; n=100)$

Likelihood: $\mathcal{L}(65|p= 0.5 \; \& \; n=100)$

Likelihood: $\mathcal{L}(65|p= 0.7 \; \& \; n=100)$

```{r}
dbinom(65,100,0.3)  
dbinom(65,100,0.5)     
dbinom(65,100,0.7)
```

## In other words

Likelihood: $\mathcal{L}(65|p= 0.3 \; \& \; n=100)$

Likelihood: $\mathcal{L}(65|p= 0.5 \; \& \; n=100)$

Likelihood: $\mathcal{L}(65|p= 0.7 \; \& \; n=100)$

```{r}
dbinom(65,100,0.3)  
dbinom(65,100,0.5)     
dbinom(65,100,0.7)
```

```{r}
dbinom(65,100,seq(0.01,0.99, by=0.01))
```

## Likelihood

```{r}
x=seq(0.01,0.99, by=0.01)
y=dbinom(65,100,seq(0.01,0.99, by=0.01))
plot(x,y,type="l", col="blue", lwd=2,ylab="Likelihood", xlab="Probability")
abline(v=0.5)

```

"Very Unlikely that it is fair"

## Probability vs Likelihood

Probability:

```{r eval=FALSE}
dbinom(0:100,100,0.5)
```

Likelihood

```{r eval=FALSE}
dbinom(65,100,seq(0.01,0.99, by=0.01))
```

```{r echo=FALSE}
x=seq(0.01,0.99, by=0.01)
y=dbinom(65,100,seq(0.01,0.99, by=0.01))
plot(x,y,type="l", col="blue", lwd=2,ylab="Likelihood", xlab="Probability")
```

Likelihood is continuous

## Review

-   If you capture 5 raccoons with weights of 11.2, 12, 11.5, 12.8, 11.06 is it more likely that the population mean is 11.5 or 17?

-   If you get 80 heads in 100 tosses, is it more likely that p is 0.8 or 0.5

-   If you count insects per tree and find 77, 70, 75, 86, and 90 is it more likely that $\lambda$ for insects per tree is 60, 80, or 120?

-   Likelihood is estimated!

-   $\mathcal{L}(x|\theta = y)$

-   And usually (and correctly): $\mathcal{L}(\theta|data)$

-   We use the probability density functions

## Likelihood and probability (correct way)

$$
\Huge \mathcal{L}(\theta|data)
$$

$$
\Huge P(data|\theta)
$$

$$ 
\mathcal{L}(\theta|data) = P(data|\theta)
$$

-   Likelihood of p being 0.5 given 48 heads out of 100 tosses

```{r}
dbinom(65,100,0.5)  

```

-   Probability of 48 heads out of 100 tosses given a probability of 0.5

```{r}
dbinom(65,100,0.5)  

```

## Bayesian

A different statistical framework

We have used frequentist so far

| Topic | Frequentist | Bayesian |
|----|----|----|
| Parameters | Fixed unknown | Random variable |
| Data | Random | Fixed |
| Inference | Based on a sampling distribution | Based on posterior |
| Output | P-value, Confidence intervals, betas | Credible intervals and posterior distribution |
| Computational needs | low | high |

-   Data, random? Yes...

-   $y_i = \beta_0 + \beta_1 + \epsilon_i$

## Bayesian take away

Different way of looking at things... we will explore how

## Bayes Theorem

$$
\Huge P(\theta |data) = \frac{P(data|\theta) \times P(\theta)}{P(data)}
$$

-   $P(\theta|data)$ : Posterior, probability of theta, given the data
-   $P(data|\theta)$ : Likelihood! $\mathcal{L}(\theta|data) = P(data|\theta)$
-   $P(\theta)$: Prior (who knows what this is!)
-   $P(data)$: Marginal (who knows!)

## Challenges

-   Challenge 1:

    -   You go out to a Forest that has been separated in multiple plots. You are wondering what the count of total trees per plot is ($\lambda$). You only can count the number of trees in one plot chosen at random. You count 37 trees, Estimate the Likelihood of different values of $\lambda$ as well as the maximum likelihood.

-   Challenge 2:

    -   You go to the same Forest, but were able to count three plots. They had 37, 41 and 35 trees. Estimate the Likelihood of different values of $\lambda$
