---
title: "Bayesian statistics: part three"
author: "SNR 610"
execute: 
  echo: true
format:
  revealjs:
    incremental: true   
    scrollable: true
    smaller: true
    chalkboard: true
    mermaid:
      theme: default
    fontsize: 55px
editor: visual 
---

## Bayesian

-   Gives us a posterior, based on a prior an a Likelihood --\> models a probability. It is obtained by the product of:

-   Likelihood, probability of a parameter being something, given our data

-   Prior: the information we had about a system

## Challenge. Bats per net

-   You are surveying bats near a beach. You set up one mist net and caught 8 bats

-   From previous surveys, you know that the mean catch per net is around 10 bats, with a standard deviation of around 3 bats

-   Estimate Likelihood, prior, combine them (joint) and estimate the posterior (normalized so that it sums to 1)

-   Plot all three on the same graph: likelihood, prior, and posterior --\> Scale the prior and likelihood if necessary so they fit on the same plot

-   Find the most probable value of $\lambda$

## Step1: Likelihood

What would the first step be?

```{r}
Lambda=seq(0,22, by=0.1)
Likelihood=dpois(8,Lambda)
plot(Lambda,Likelihood,type="l", col="blue", lwd=2,ylab="Likelihood", xlab="Lambda")
```

## Step 2: Prior {.nostretch}

```{r}
#| fig-height: 4
#| fig-width: 6
#| fig-opts: fig.align='center'
#| out-width: "80%"

mu <- 10
sigma <- 3

beta <- mu / sigma^2
alpha <- mu * beta

prior <- dgamma(Lambda, shape = alpha, rate = beta)


plot(Lambda,prior,type="l", col="salmon", lwd=2,ylab="Likelihood", xlab="Lambda")
lines(Lambda,Likelihood,, col="blue", lwd=2)
```

## Step 3: Joint

```{r}
jointL<-prior*Likelihood
```

## Step 4: Posterior and plot

```{r}
Posterior<- jointL/sum(jointL)
plot(Lambda,Posterior,type="l", lwd=4, lty=2,ylab="joint Likelihood", xlab="Lambda")

```

## Plot them together

```{r}
plot(Lambda,Posterior,type="l", lwd=4, lty=2,ylab="joint Likelihood", xlab="Lambda")
lines(Lambda,Likelihood, col="blue", lwd=2)
lines(Lambda,prior, col="salmon", lwd=2)
```

## Plot them again {.nostretch}

Need scaling

Based on posterior

```{r}
#| fig-height: 4
#| fig-width: 6
#| fig-opts: fig.align='center'
#| out-width: "80%"
L_scaled <- Likelihood * max(Posterior) / max(Likelihood)
prior_scaled <- prior * max(Posterior) / max(prior)


plot(Lambda,Posterior,type="l", lwd=4, lty=2,ylab="joint Likelihood", xlab="Lambda")
lines(Lambda,L_scaled, col="blue", lwd=2)
lines(Lambda,prior_scaled, col="salmon", lwd=2)
```

## How do humans think

```{mermaid}
%%| echo: false
flowchart LR
  A[Experience] --> B[Observation]
  B --> C[Updated Belief]
  C --> A



```

Be it human relations, weather, or sports!

-   We do not think about about things in "point estimates" but more distribution like

## Comparing methods (Bayesian vs Frequentist)

Assume you have data for 37 night nets, rather than just one. Same prior as before (mean =10 and sd =3)

How would we estimate the posterior?

```{r echo=FALSE}
bats <- rpois (37, lambda = 9.13)
```

```{r}
print(bats)
```

## Bayesian {.nostretch}

```{r}
#| fig-height: 4
#| fig-width: 6
#| fig-opts: fig.align='center'
#| out-width: "80%"

Lambda=seq(0,22, by=0.1)

Likelihood=sapply(Lambda, function(l) prod(dpois(bats, l)))

mu <- 10
sigma <- 3

beta <- mu / sigma^2
alpha <- mu * beta

prior <- dgamma(Lambda, shape = alpha, rate = beta)

jointL<-prior*Likelihood

Posterior<- jointL/sum(jointL)
plot(Lambda,Posterior,type="l", lwd=4, lty=2,ylab="joint Likelihood", xlab="Lambda")

```

## Frequentist

```{r}
mean(bats)
```

-   Which one is harder?

## Different outputs

#### Frequentist

One estimate (most likely value; expected value)

Uncertainty is given by Confidence Intervals

#### Bayesian

A posterior distribution

Represents our updated belief about $\lambda$

Uncertainty is both shape and spread!

## How about two parameters?

-   Imagine now you have continuous data

-   You have a mean

-   You have variance

-   Bayesian needs a 2-d likelihood surface

## What is a 2-d surface

Imagine you want to test the likelihood of $\mu$ from 70 to 80

You want to tests the likelihood of $\sigma$ from 3 to 7

Each time you run a likelihood estimate you run a joint likelihood of both mean and sd and you want to test all potential combinations!

For a **normal model** with data

$$y_i \sim \text{Normal}(\mu, \sigma) $$

We need to explore both parameters:

$$P(\mu, \sigma \mid y) \propto P(y \mid \mu, \sigma) \times P(\mu, \sigma) $$

**Frequentist:**\
- Closed-form solutions\
$$\hat{\mu} = \bar{y} \\  \hat{\sigma} = \text{sd}(\bar{y})$$

**Bayesian:**\
- Must compute (or sample) over the entire **2D surface**!

------------------------------------------------------------------------

```{r}
set.seed(123)
y <- rnorm(20, mean = 75, sd = sqrt(22))

# Define grid for mu and sigma
mu_vals <- seq(70, 80, by=0.01 )
sigma_vals <- seq(3, 7, by=0.01)

# Compute log-likelihood surface
ll <- outer(mu_vals, sigma_vals, Vectorize(function(mu, sigma) {
  sum(dnorm(y, mean = mu, sd = sigma, log = TRUE))
}))

# Convert to likelihood
lik <- exp(ll - max(ll))  # rescale for plotting

```

## Plot

```{r}

# Contour plot of the 2D likelihood surface

contour(mu_vals, sigma_vals, lik,
xlab = expression(mu), ylab = expression(sigma),
main = "2D Likelihood Surface")
points(mean(y), sd(y), col = "red", pch = 19)
text(mean(y), sd(y), "MLE", pos = 4, col = "red")


```

## Takeaway

-   In Bayesian, if we are doing one estimate ($\lambda$) we can do a curve

-   If we are doing two estimates, we need a contour plot, or a surface plot

-   For two estimates we have essentially a matrix

-   More than two estimates?

## Example

Example from assignment 5

```{r echo=FALSE}

cod<-read.csv("../data/parasitecod.csv")
library(dplyr)
cod<-cod%>%mutate(across(c(Year,Sex,Stage,Area),as.factor))

```

```{r}
codmodel1<-glm(Prevalence~Length+Year+Area,data=cod,family = binomial(link="logit"))
summary(codmodel1)

```

-   7 different estimates!

-   Joint Likelihood for each one!

-   Really hard to choose a starting value!

-   If we get 8 bats, easy to choose a beta. In a model it is hard to choose a beta!

## Many parameters

-   Imagine a simple linear model

-   $$ y_i \sim \beta_0 + \beta_1x_{1i} +  \beta_2x_{2i} + \beta_3x_{3i} + \epsilon_i $$

-   Posterior: $P (\beta_0, \beta_1, \beta_2, \sigma | data)$

-   We cannot make a grid! We cannot visualize!

## How do models solve this?

In frequentist: Equations that minimize residuals. It uses calculus and algebra

-   You get the same solution each time. If all of you run the same model with the same data in R, you will all get the same betas as me!

-   Bayesian Integrates a posterior --\> No closed solution!

-   Different solution each time!

-   Why? Bayesian cannot solve this using an equation. So, it samples from it using Markov Chain Monte Carlo (MCMC!) and obtains a probability from it

## MCMC?

-   Ever heard of this?

-   Let's try this! With a game!

## In-Pairs Exercise: Acting Out an MCMC Random Walk

### Goal

Experience how MCMC explores a likelihood surface step-by-step. NO COMPUTER NEEDED!

DO NOT RUN ANYTHING YET!

Choose one student to do the random walk, one to be the computer or tester

## Exercise

### Goal

Experience how MCMC explores a likelihood surface step-by-step. NO COMPUTER NEEDED!

------------------------------------------------------------------------

### Setup

1.  **Student A:** open and run the following code (DO NOT SHOW TO STUDENT B!):

```{r eval=FALSE}
x <- seq(-50, 150, by = 1)

# Create a landscape with 3 peaks of different widths and heights
likelihood <- 
  0.7 * dnorm(x, mean = 20, sd = 10) +
  1.0 * dnorm(x, mean = 80, sd = 15) +
  0.4 * dnorm(x, mean = 120, sd = 5)

# Normalize so it looks like a probability surface
likelihood <- likelihood / max(likelihood)

plot(x, likelihood, type = "l", lwd = 3, col = "darkgreen",
     xlab = "Parameter value (x)",
     ylab = "Relative likelihood",
     main = "Example Likelihood Landscape")


```

2.  **Student B:** secretly pick a **starting point** (between 0 and 100). Call this x

### The Walk

1.  **Student B:** Write down the starting point (x) and mention it to student A. Student A tell student B an approximate value of the Likelihood. Write it down.

2.  **Student B:** flips a “coin” in R:

```{r}
rbinom(1, 1, 0.5)
```

If 1 --\> propose x + 3

If 0 --\> propose x – 3

3.  **Student A:** checks the new height (likelihood) and compares it to the current. If new height \> current, move 3 steps in that direction (Student B: write down the new x, student A write Likelihood) If not, move only 1 step. Write down the new x and Likelihood as well.

Repeat \~20 times

Plot in paper (x value vs Likelihood)

## MCMC

Start with a random guess.

Propose a new set of parameters.

Accept or reject based on how likely they are given the data.

Repeat thousands of times. The chain “learns” the posterior shape.

## Youtube video

## Gibbs sampler

Most of our software uses a Gibbs sampler.

-   It tests one parameter at a time.

-   Fixes some parameters and moves one around

-   After a while, it switches to a different parameter, and so on

## Next classes (suggestion)

-   Wednesday: Plots

-   Friday: A Bayesian example (only for those interested).

-   Next Week: Multivariate (lecture). Friday lab, no need to turn in, but an exercise I want you to do

-   Monday 24: Session to work on final project (no class)

-   Monday 1st. Session to work on final project (no class). Free coffee + snacks (RSVP)!
